Optimizing a PoS Tag Set for Dependency Parsing

### 2016-05-06 submission.tex
changes to tag set experiments on nouns

\subsection{Tag Set Experiments}
We modified the tags for nouns (\texttt{subst}), verbs (\texttt{verb}),
adjectives (\texttt{adj}), determiners (\texttt{det}) and pronouns
(\texttt{pron}) in NDT by appending selected sets of morphological features to
each tag in order to increase the linguistic information expressed by the tags.
For each tag, we first experimented with each of the available features in
isolation before employing various combinations of them. We base our choices
of combinations on how promising the features are and what we deem worth
investigating in terms of linguistic utility, in order to see how the features
might interact.

The morphological properties of the various parts-of-speech are reflected in
the morphological features associated with the respective PoS tags. For
instance, as nouns in Norwegian can take on gender, definiteness and number,
additionally marked with genitive case, the treebank operates with features for
gender, definiteness, number and case.
%with "accompanying" values
In addition to morphological properties such as definiteness, tense and number,
all classes except for verbs have a \emph{type} feature that provides
information about the subtype, e.g., common nouns and proper nouns, while
pronouns can be reflexive, reciprocal, personal or interrogative.

%"Nearly" all tag set modifications leads to a drop in tagger accuracy, as
%expected. Few exceptions... We observed increases in parser accuracy scores
%with linguistically motivated tag set modifications for all PoS tags we
%experimented with, the most improved being noun with an increase in LAS from
%the original tag set by 1.80 percentage points.

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine or neuter),
definiteness (definite or indefinite) and number (singular or plural). There
is agreement in gender, definiteness and number between nouns and their
modifiers (adjectives and determiners). Additionally, NDT has a separate case
feature for distinguishing nouns in genitive case.

As nouns constitute the largest class in the treebank by far, the effects are
greatest for noun tokens in terms of overall change in performance. The results
of experimenting with tag set modifications for nouns are shown in Table
\ref{substresults}, and reveal that, apart from case, none of the tag set
modifications improves the tagging. However, they all give rise to increases in
parser accuracy scores. Genitive case marks possession, hence nouns marked with
genitive case are quite different from other nouns, taking a noun phrase as
complement. Distinguishing on type is useful and informative, as evident by the
presence of separate tags for proper and common nouns in many tag sets, such as
those of Penn Treebank and Stockholm-Umeå corpus.

The most informative features are definiteness, which
leads to an increase in LAS by 1.26 percentage points to 88.27\%, and type,
yielding an LAS of 88.07\%. Turning to combinations of features, we found that
the combination of type and case, as well as type and definiteness, were the
most promising, which led us to combine type, case and definiteness in a final
experiment, resulting in LAS of 88.81\% and UAS of 91.73\%, constituting large
increases from parsing with the original tag set, by 1.80 percentage points
and 1.54 percentage points, respectively.

@@

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite, past perfect) in
Norwegian and additionally exhibit mood (imperative, indicative) and voice
(active or passive). Note that both voice and mood have only a single value in
the treebank, \texttt{pass} (passive) and \texttt{imp} (imperative),
respectively.  Verbs which are not passive are implicitly active, and verbs
which are not imperative are indicative.

>>>>

\subsection{Tag Set Experiments}
We modify the tags for nouns (\texttt{subst}), verbs (\texttt{verb}),
adjectives (\texttt{adj}), determiners (\texttt{det}) and pronouns
(\texttt{pron}) in NDT by appending selected sets of morphological features to
each tag in order to increase the linguistic information expressed by the tags.
For each tag, we first experiment with each of the available features in
isolation before employing various combinations of them. We base our choices of
combinations on how promising the features are and what we deem worth
investigating in terms of linguistic utility, in order to see how the features
might interact.

The morphological properties of the various parts-of-speech are reflected in
the morphological features associated with the respective PoS tags. For
instance, as nouns in Norwegian can take on gender, definiteness and number,
the treebank operates with features for gender, definiteness and number
complementing the \texttt{subst} tag. In addition to morphological properties
such as definiteness, tense and number, all classes except for verbs have a
\emph{type} feature that provides information about the subtype of the PoS,
e.g., whether a noun is common or proper.

\paragraph{Nouns}
In Norwegian, nouns are assigned gender (feminine, masculine or neuter),
definiteness (definite or indefinite) and number (singular or plural). There is
agreement in gender, definiteness and number between nouns and their modifiers
(adjectives and determiners). Additionally, NDT has a separate \emph{case}
feature for distinguishing nouns in genitive case. Genitive case marks
possession, hence nouns marked with genitive case are quite different from
other nouns, taking a noun phrase as complement. Distinguishing on type is
useful and informative, as evident by the presence of separate tags for proper
and common nouns in many tag sets, such as those of Penn Treebank and
Stockholm-Umeå corpus.

%As nouns constitute the largest class in the treebank by far, the effects of
%the tag set modifications are greatest for noun tokens in terms of overall
%change in performance.
The results of experimenting with tag set modifications
for nouns are shown in Table \ref{substresults} and reveal that, apart from
case, none of the tag set modifications improve the tagging. However, they all
give rise to increases in parser accuracy scores.
The most informative features are definiteness, which
leads to an increase in LAS by 1.26 percentage points to 88.27\%, and type,
yielding an LAS of 88.07\%. Turning to combinations of features, we found that
the combination of type and case, as well as type and definiteness, were the
most promising, which led us to combine type, case and definiteness in a final
experiment, resulting in LAS of 88.81\% and UAS of 91.73\%, constituting large
improvements from parsing with the original tag set, of 1.80 percentage points
and 1.54 percentage points, respectively.

@@

\paragraph{Verbs}
Verbs are inflected for tense (infinitive, present, preterite or past perfect)
in Norwegian and additionally exhibit mood (imperative or indicative; also
conjunctive, however, it is very uncommon and does not occur in the treebank)
and voice (active or passive). Note that both voice and mood have only a single
value in the treebank; \texttt{pass} (passive) and \texttt{imp} (imperative),
respectively.  Verbs which are not passive are implicitly active, and verbs
which are not imperative are indicative.
